Verifying Interactive WebSocket Applications in TypeScript

### 2020-01-24 PLACES_2020/main.tex
More FZ comments, latex file formatting

\usepackage{listings}
\lstdefinelanguage{Scribble}{
  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
  keywordstyle=\color{blue}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{darkgray}\ttfamily,
  stringstyle=\color{purple}\ttfamily,
  morestring=[b]',
  morestring=[b]",
}
\lstdefinelanguage{JavaScript}{
  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
  keywordstyle=\color{blue}\bfseries,
%  ndkeywords={class, export, boolean, throw, implements, import, this},
%  ndkeywordstyle=\color{darkgray}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{darkgray}\ttfamily,
  stringstyle=\color{purple}\ttfamily,
  morestring=[b]',
  morestring=[b]"
}
\lstset{
	basicstyle=\footnotesize\ttfamily,
	numbers=left,
	frame=single,
	tabsize=4
}
\usepackage{caption}

@@

Multiparty session types (MPST) provide a framework for verifying conformance
of implementations to their prescribed communication protocol. Existing
proposals for applying the MPST framework in application development either
neglect the event-driven nature of web applications, or lack compatibility with
industry tools and practices which discourages mainstream adoption by web developers.

In this paper, we present an implementation of the MPST framework for developing interactive web applications using familiar industry tools in TypeScript and the \textit{React.js} framework. The global communication protocol is specified in the Scribble protocol language and checked for well-formedness by the Scribble toolchain. We obtain the local protocol for each role, which describes the interactions of the global communication protocols observed by the role. We encode the local protocol into TypeScript types, catering for server-side and client-side targets separately, and show that our encoding guarantees that only implementations which conform to the protocol will type-check. We demonstrate the effectiveness of our approach through a web-based implementation of the classic \textit{Noughts and Crosses} game from a MPST formalism of the game logic.

@@

Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers.
Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as real-time multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
Enabling more complex communication patterns caters for more interactive use cases, but introduces additional correctness concerns to the developer.

% Example: noughts and crosses
Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players. Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied). A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}. SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser. Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.

@@

\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect. By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.

@@

Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble \cite{Scribble}, a MPST-based protocol specification language and code generation toolchain. We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.

@@

global protocol Game(role Svr, role P1, role P2) {
  Pos(Point) from P1 to Svr;
  choice at Svr {
    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
  } or {
    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
  } or {
    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
    do Game(Svr, P2, P1);
  }
}
\end{lstlisting}
\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
\label{lst:game}
\end{figure}

We leverage the Scribble toolchain to check for protocol
well-formedness.
This directly corresponds to multiparty session
type theory \cite{MPST}:
\fz{Cite the featherweight scribble paper for the correspondence?}
a Scribble protocol maps to some \textit{global type}, and protocol verification implements the algorithmic projection
defined in \cite{MPST} to derive valid local type \textit{projections} for all participants.  We obtain a set of \textit{endpoint protocols}
-- one for each role from a well-formed global protocol. An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain. We use the EFSMs as a basis for API generation and adopt the formalisms in \cite{Hybrid2016}.

@@

Developers can implement their application using APIs generated from the EFSM to guarantee correctness by construction.
Our approach integrates the EFSM into the development workflow by encoding
session types as TypeScript types.
There are inherent differences between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser. This motivates our design choice of generating different session types encodings for server (\sectionref{section:server}) and browser targets (\sectionref{section:browser}).

@@

\begin{wrapfigure}{r}{0.5\textwidth}
  \begin{center}
    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
  \end{center}

  \captionof{figure}{EFSM for \texttt{Svr}.}
  \label{fig:efsmsvr}
\end{wrapfigure}

We refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
running example in this section. For server-side targets, we encode
EFSM states as types, considering
receive and send states separately. Each TypeScript encoding is
assigned to its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state. For any state $S$ in the EFSM, we refer to the TypeScript type alias of its encoding as $\llbracket S \rrbracket$.

We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide \textit{static} linearity guarantees (\sectionref{section:serverlinear}). Our encoding sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encoding does not concern with the role involved in the send/receive action. We outline the encoding below using examples from the \textit{Noughts and Crosses} game (Listing \ref{lst:svr}).

\paragraph{Branching state} We consider a receive state as a unary branching state for conciseness. A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all branches, corresponding to a member field. A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}. The developer implements a branching operation by passing callbacks for each branch, parameterised by the expected message payload type for that branch.

@@

\begin{figure}[!h]
\begin{lstlisting}[language=JavaScript]
export type S13 = { Pos: (payload: Point) => S15 };
export type S15 = [Labels.Lose, Point, S16]
                | [Labels.Draw, Point, S17]
                | [Labels.Update, Point, S18];
\end{lstlisting}
\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
\label{lst:svr}
\end{figure}

@@

\paragraph{Repeat use} We do not expose channels to the programmer, which makes \textit{reuse} impossible. For example, to send a message, the generated API only requires the payload that needs to be sent, and the session runtime performs the send internally, guaranteeing this action is done \textit{exactly once} by construction.

@@

We refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
running example in this section. Preserving behavioural typing and
channel linearity are known to be challenging for browser-side
applications due to EFSM transitions being triggered by user events:
in the case of \textit{Noughts and Crosses}, once the user makes a move
by clicking on a cell on the game board, this click event must be
deactivated until the user's next turn, otherwise the user can click
again and violate channel linearity. Our design goal is to enforce
this statically through the generated APIs.

For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture. Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.

@@

@@

\paragraph{Components} A component is a reusable UI element which
contains its own markup and logic. Components implement a
\texttt{render()} function which returns a React element, the smallest
building blocks of a React application. Components can keep
\textit{state} and the \texttt{render()} function is invoked upon a
change of state. A simple counter can be implemented as a component,
with \texttt{count} stored as state, a button which increments \texttt{count}
when clicked and a \texttt{div} that renders the current
\texttt{count}. Components can also render other components, which gives rise
to a parent/child relationship between components. Parents can pass
data to children as \textit{props}: going back to the aforementioned
example, the the counter component could render a child component
\texttt{<StyledDiv count=\{this.state.count\} />} in its
\texttt{render()} function, propagating the \texttt{count} from its
state to the child. This enables reusability, and for our use case,
gives a control to the parent on what data to pass to its children (e.g. pass the payload of a received message to a child to render).

\fz{Is VDOM needed to be covered here?}
\paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly. This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.

@@

\paragraph{State} An EFSM state is encoded as an \textit{abstract} React component. This is an abstract class to require the programmer to provide their own view function, which translates conveniently to the \texttt{render()} function of React components. Our session runtime (\sectionref{section:clientruntime}) ``executes'' the EFSM and renders the current state, so upon transitioning to a successor state, the successor's view function will be invoked, as per the semantics expressed in \cite{MVU2019}.

\paragraph{Model transitions} Transitions are encoded as React component props onto the encoded states by the session runtime (\sectionref{section:clientruntime}). We motivate the design choice of not exposing channel resources to provide guarantees on channel usage. React components in TypeScript are \textit{generic} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for. The model transitions for EFSMs are message send and receive.

\subparagraph{Send} We make the assumption that message sending is triggered by some user-driven UI event (e.g. clicking a button, pressing a key on the keyboard) which interacts with some DOM element. We could pass a \texttt{send()} function as a prop to the sending state, but the programmer would be free to call the function multiple times which makes channel reuse possible. Instead, we pass a \textit{factory function} as a prop, which will, given an HTML event and a event handler function, return a fresh React component that binds the sending action on construction, so once the bound event is triggered, our session runtime executes the event handler function to obtain the payload to send, perform the send \textit{exactly once} and transition to (which, in practice, means render) the successor state.

@@

We clarify the semantics using the \textit{Noughts and Crosses} example in Listing \ref{lst:clientapp}. \texttt{this.props.Pos} is the factory function prop passed from the session runtime. For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.

@@

Our session runtime can be interpreted as an abstraction on top of the React VDOM that implements the EFSM by construction. The session runtime itself is a React component too, named after the endpoint role identifier: it opens the WebSocket connection to the server, keeps track of the current EFSM state as part of its React component state, and most importantly, renders the React component encoding of the active EFSM state. Channel communications are managed by the runtime, which allows it to render the successor of a receive state upon receiving a message from the channel. Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components. The session runtime component is rendered by the programmer and requires, as props, the \textit{endpoint URL} (so it can open the connection) and a list of \textit{concrete state components}.

The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components. The session runtime requires references to these concrete components order to render the user implementation accordingly.

@@

Communication channels are not exposed to the programmer so multiple sends are impossible. This does not restrict the programmer from binding the send action to exactly one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once the cell is clicked, the send is only performed once and the successor state is rendered on the DOM, so the channel resource used to send becomes unavailable.

@@

\paragraph{Game server} We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime. We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result. This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic.

Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}. There is also full type erasure when transpiling to JavaScript to run the server code, so state space explosions is never a runtime consideration.

@@

\paragraph{Session types in web development} \cite{PureScript2019} targeted web development in PureScript using the \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as multi-parameter type classes. However, it presents a trade-off between achieving static linearity guarantees from the type-level EFSM encoding under the expressive type system and providing an intuitive development experience to programmers, especially given the prevalence of JavaScript and TypeScript applications in industry. \cite{MVU2019} focused on applying binary session types in front-end web development and presented approaches that tackle the challenge of guaranteeing linearity in the event-driven environment.

@@

Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
Whilst our approach supports multiparty sessions, the nature of WebSockets require some server-based role in the communication protocol and clients to interact via the server. Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.

>>>>

\usepackage{listings}
\usepackage[noabbrev, capitalise]{cleveref}
\lstdefinelanguage{Scribble}{
  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
  keywordstyle=\color{blue}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{darkgray}\ttfamily,
  stringstyle=\color{purple}\ttfamily,
  morestring=[b]',
  morestring=[b]",
}
\lstdefinelanguage{JavaScript}{
  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
  keywordstyle=\color{blue}\bfseries,
%  ndkeywords={class, export, boolean, throw, implements, import, this},
%  ndkeywordstyle=\color{darkgray}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{darkgray}\ttfamily,
  stringstyle=\color{purple}\ttfamily,
  morestring=[b]',
  morestring=[b]"
}
\lstset{
	basicstyle=\footnotesize\ttfamily,
	numbers=left,
	frame=single,
	tabsize=4
}
\usepackage{caption}

@@

Multiparty session types (MPST) provide a framework for verifying conformance
of implementations to their prescribed communication protocol.
Existing proposals for applying the MPST framework in application development
either neglect the event-driven nature of web applications, or lack
compatibility with industry tools and practices, which discourages mainstream
adoption by web developers.

In this paper, we present an implementation of the MPST framework for
developing interactive web applications using familiar industry tools in
TypeScript and the \textit{React.js} framework.
The developer can use the Scribble protocol
language to specify the protocol and use the Scribble toolchain to validate
and obtain the local protocol for each role.
The local protocol describes the interactions of the global communication
protocols observed by the role.
We encode the local protocol into TypeScript types, catering for server-side
and client-side targets separately, and show that our encoding guarantees that
only implementations which conform to the protocol can type-check.
We demonstrate the effectiveness of our approach through a web-based
implementation of the classic \textit{Noughts and Crosses} game from a MPST
formalism of the game logic.

@@

Modern interactive web applications aim to provide a highly responsive user
experience by minimising the communication latency between clients and servers.
Whilst the HTTP request-response model is sufficient for retrieving static
assets, applying the same stateless communication approach for interactive use
cases (such as real-time multiplayer games) introduces undesirable performance
overhead.
Developers have since adopted other communication transport abstractions over
HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy
low-latency full-duplex client-server communication in their applications over
a single persistent connection.
Enabling more complex communication patterns caters for more interactive use
cases, but introduces additional correctness concerns to the developer.

% Example: noughts and crosses
Consider a classic turn-based board game of \textit{Noughts and Crosses}
between two players.
Both players, identified by either noughts or crosses respectively, take
turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
wins (when their markers form one straight line on the board) or a stalemate is
reached (when all cells are occupied).
A web-based implementation may involve players connected to a game server via
WebSocket connections.
The players interact with the game from their web browser, which shows a
\textit{single-page application} (SPA) of the game client written in a popular
framework like \textit{React.js} \cite{React}.
SPAs feature a single HTML page and dynamically rendered content via JavaScript
in the browser.
Players take turns to make a move on the game board, which sends a message to
the server, and the server implements the game logic to progress the game
forward until a result (either a win/loss or draw) can be declared, where
either the move of the other player or the game result is sent to players.

@@

\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for
formally specifying a structured communication pattern between concurrent
processes and verifying implementations for correctness with respect to the
communications aspect.
By specifying the client-server interactions of our game as an MPST protocol
and verifying the implementations against the protocol for well-formedness,
MPST theory guarantees well-formed implementations to be free from
communication errors.

@@

Development begins by specifying the permitted communications between
participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
MPST-based protocol specification language and code generation toolchain.
We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
\cref{lst:game}.

@@

global protocol Game(role Svr, role P1, role P2) {
  Pos(Point) from P1 to Svr;
  choice at Svr {
    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
  } or {
    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
  } or {
    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
    do Game(Svr, P2, P1); // Continue the game with player roles swapped
  }
}
\end{lstlisting}
\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
\label{lst:game}
\end{figure}

We leverage the Scribble toolchain to check for protocol
well-formedness.
This directly corresponds to multiparty session
type theory \cite{MPST}:
\fz{Cite the featherweight scribble paper for the correspondence?}
a Scribble protocol maps to some \textit{global type}, and protocol
verification implements the algorithmic projection defined in \cite{MPST} to
derive valid local type \textit{projections} for all participants.
% Note: Scribble Projection is not MPST projection, because you can have
% protocols with choices that send to different roles.
We obtain a set of \textit{endpoint protocols} -- one for each role from a
well-formed global protocol.
An endpoint protocol only preserves the interactions defined by the global
protocol in which the target role is involved, and corresponds to an equivalent
\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
We use the EFSMs as a basis for API generation and adopt the formalisms in
\cite{Hybrid2016}.

@@

Developers can implement their application using APIs generated from the EFSM
to guarantee correctness by construction.
Our approach integrates the EFSM into the development workflow by encoding
session types as TypeScript types.
There are inherent differences between front- and back-end web development:
business logic tends to initiate communication in the back-end, whilst
communication in the front-end is mostly driven by user interactions with the
browser. This motivates our design choice of generating different session types
encodings for server (\cref{section:server}) and browser targets
(\cref{section:browser}).

@@

\begin{wrapfigure}{r}{0.5\textwidth}
  \vspace{-5mm}
  \begin{center}
    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
  \end{center}

  \vspace{-5mm}
  \captionof{figure}{EFSM for \texttt{Svr}.}
  \label{fig:efsmsvr}
\vspace{-1cm}
\end{wrapfigure}

We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
this section.
For server-side targets, we encode EFSM states into TypeScript types and
considering receive and send states separately.
Each TypeScript encoding is assigned to its state identifier from the EFSM as
its type alias, thus providing syntactic sugar when referring to the successor
state in the TypeScript encoding of the current state.
For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
encoding as $\llbracket S \rrbracket$.

\paragraph{Branching state}
We consider a receive state as a unary branching state for conciseness.
A branching state is encoded as an \textit{object literal}
\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
branches, corresponding to a member field.
A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
\textit{member field} named $\texttt{label}_i$ of function type
$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
The developer implements a branching operation by passing callbacks for each
branch, parameterised by the expected message payload type for that branch.

@@

\begin{figure}[ht]
\begin{lstlisting}[language=JavaScript]
export type S13 = { Pos: (payload: Point) => S15 };
export type S15 = [ Labels.Lose, Point, S16 ]
                | [ Labels.Draw, Point, S17 ]
                | [ Labels.Update, Point, S18 ];
\end{lstlisting}
\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
\label{lst:svr}
\end{figure}

We make a key design decision \textit{not} to expose communication channels in
the TypeScript session type encoding to provide \textit{static} linearity
guarantees (\cref{section:serverlinear}).
Our encoding sufficiently exposes seams for the developer to inject their
business logic, whilst the generated session API
(\cref{section:serversessionapi}) handles the sending and receiving of
messages;
as a result, the encoding does not concern with the role involved in the
send/receive action. We outline the encoding below using examples from the
\textit{Noughts and Crosses} game (\cref{lst:svr}).

@@

\paragraph{Repeated use}
We do not expose channels to the developer, which makes \textit{reuse}
impossible.
For example, to send a message, the generated API only requires the payload
that needs to be sent, and the session runtime performs the send internally,
guaranteeing this action is done \textit{exactly once} by construction.

@@

@@

We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
this section.
Preserving behavioural typing and channel linearity are known to be challenging
for browser-side applications due to EFSM transitions being triggered by user
events:
in the case of \textit{Noughts and Crosses}, once the user makes a move by
clicking on a cell on the game board, this click event must be deactivated
until the user's next turn, otherwise the user can click again and violate
channel linearity.
Our design goal is to enforce this statically through the generated APIs.

For browser-side targets, we extend the approach presented in \cite{MVU2019} on
\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
architecture.
Each state in the EFSM is a \emph{model type} and uniquely defines a
\textit{view function}, set of \textit{messages} and \textit{update function}:
the view function defines what to render to the DOM, the set of messages define
the possible (IO) actions available at that state, and the update function
defines which successor state to transition to, given some supported IO action
at this state.

@@

\paragraph{Components}
A component is a reusable UI element which
contains its own markup and logic.
Components implement a \texttt{render()} function which returns a React
element, the smallest building blocks of a React application, analogous to the
view function in the MVU architecture.
Components can keep \textit{state} and the \texttt{render()} function is
invoked upon a change of state.
A simple counter can be implemented as a component,
with \texttt{count} stored as state, a button which increments \texttt{count}
when clicked and a \texttt{div} that renders the current
\texttt{count}.
Components can also render other components, which gives rise
to a parent/child relationship between components.
Parents can pass data to children as \textit{props} (short for properties).
Going back to the aforementioned example, the the counter component could
render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
its \texttt{render()} function, propagating the \texttt{count} from its state
to the child.
This enables reusability, and for our use case, gives a control to the parent
on what data to pass to its children (e.g. pass the payload of a received
message to a child to render).

\fz{Is VDOM needed to be covered here?}
\paragraph{Virtual DOM (VDOM)}
React components are rendered on a logical
abstraction of the DOM, which in turn performs a \texttt{diff} on the current
browser DOM and patches the delta accordingly.
This allows the developer to
directly specify what should be rendered without having to worry about which
elements to append or remove from the browser DOM.

@@

\paragraph{State}
An EFSM state is encoded as an \textit{abstract} React
component.
This is an abstract class to require the developer to provide their
own view function, which translates conveniently to the \texttt{render()}
function of React components.
Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
renders the current state.
Upon transitioning to a successor state, the successor's view function will be
invoked, as per the semantics expressed in \cite{MVU2019}.

\paragraph{Model transitions}
Transitions are encoded as React component props onto the encoded states by the
session runtime (\cref{section:clientruntime}).
We motivate the design choice of not exposing channel resources to provide
guarantees on channel usage.
React components in TypeScript are
\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
types of prop and state.
The parameters allow us to leverage the TypeScript compiler to
verify that the props for model transitions stay local to the state they are
defined for.
The model transitions for EFSMs are message send and receive.

\subparagraph{Send}
We make the assumption that message sending is triggered by
some user-driven UI event (e.g. clicking a button, pressing a key on the
keyboard) which interacts with some DOM element.
We could pass a
\texttt{send()} function as a prop to the sending state, but the developer
would be free to call the function multiple times which makes channel reuse
possible.
Instead, we pass a \textit{factory function} as a prop, which will,
given an HTML event and a event handler function, return a fresh React
component that binds the sending action on construction.
So once the bound event is triggered, our session runtime executes the event
handler function to obtain the payload to send, perform the send
\textit{exactly once} and transition to (which, in practice, means render) the
successor state.

@@

We demonstrate the semantics using the \textit{Noughts and Crosses} example in
\cref{lst:clientapp}.
\texttt{this.props.Pos} is the factory function prop
passed from the session runtime.
For each x-y coordinate on the game board, we
create a \texttt{SelectPoint} React component from the factory function (which
reads ``build a React component that sends the \texttt{Pos} message with x-y
coordinates as payload when the user clicks on it'') and we wrap a table cell
(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
component to bind the click event on the table cell.

@@

Our session runtime can be interpreted as an abstraction on top of the React
VDOM that implements the EFSM by construction.
The session runtime itself is a React component too, named after the endpoint
role identifier:
it opens the WebSocket connection to the server, keeps track of the current
EFSM state as part of its React component state, and most importantly, renders
the React component encoding of the active EFSM state.
Channel communications are managed by the runtime, which allows it to render
the successor of a receive state upon receiving a message from the channel.
Similarly, the session runtime is responsible for passing the required props
for model transitions to EFSM state React components.
The session runtime component is rendered by the developer and requires, as
props, the \textit{endpoint URL} (so it can open the connection) and a list of
\textit{concrete state components}.

The developer writes their own implementation of each state (mainly to
customise how the state is rendered and inject business logic into state
transitions) by extending the abstract React class components.
The session runtime requires references to these concrete components order to
render the user implementation accordingly.

@@

Communication channels are not exposed to the developer so multiple sends are
impossible.
This does not restrict the developer from binding the send action to exactly
one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
send action to each unoccupied cell on the game board, but the generated
runtime ensures that, once the cell is clicked, the send is only performed once
and the successor state is rendered on the DOM, so the channel resource used to
send becomes unavailable.

@@

\paragraph{Game server}
We set up the WebSocket server inside an Express.js \cite{ExpressJS}
application on top of a Node.js \cite{NodeJS} runtime.
We define our own game logic in a \texttt{Board} class to keep track of the
game state and expose methods to query the result.
This custom logic is integrated into our \texttt{Receive\_P1} and
\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
can handle \texttt{Pos(Point)} messages from players and transition according
to the injected game logic.

\fz{Where is Line 6?}
Note that, by the structural typing nature of TypeScript, replacing
\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
type-correct -- this allows for better code reuse as supposed to defining
additional abstractions to work around the limitations of nominal typing in
\cite{Hybrid2016}.
There is also full type erasure when transpiling to JavaScript to run the
server code, so state space explosions is never a runtime consideration.

@@

\paragraph{Session types in web development}
\cite{PureScript2019} targeted web development in PureScript using the
\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
multi-parameter type classes.
However, it presents a trade-off between achieving static linearity guarantees
from the type-level EFSM encoding under the expressive type system and
providing an intuitive development experience to developers, especially given
the prevalence of JavaScript and TypeScript applications in industry.
\cite{MVU2019} focused on applying binary session types in front-end web
development and presented approaches that tackle the challenge of guaranteeing
linearity in the event-driven environment.

@@

Future works include incorporating \textit{explicit connection actions}
introduced in \cite{ExplicitConnections} in our API generation to better model
real-world communication protocols that may feature in interactive web
applications.
Whilst our approach supports multiparty sessions, the nature of
WebSockets require some server-based role in the communication protocol and
clients to interact via the server.
Extending support to WebRTC \fz{cite it or we don't know what it is} would
cater for peer-to-peer communication between browsers, which further opens up
possibilities for communication protocols supported by our approach.
